Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-07 tex/main.tex
more

\begin{abstract}
Model finding in SMT solvers is
generally limited to universally quantified formulas in some restricted fragment.
This paper introduces a translation which reduces axioms specifying a large
class of recursive functions, including terminating and
productive functions to universally quantified formulas for which known
techniques are applicable.
An empirical evaluation confirms that the approach improves 
the performance of existing solvers on benchmarks from two sources. 
The translation is implemented as a preprocessor in the solver CVC4.
\end{abstract}

@@

In this paper, we present a method for translating formulas involving recursice function
definitions into formulas where finite model finding techniques can be applied.
The recursive functions must meet a semantic criterion to be admissible
(Section~\ref{sec:prelim}). This criterion is met by terminating recursive
function definitions, but also by
\relax{productive} corecursive functions. It is, however, not met by inconsistent
definitions such as $\forall n : {\typeint}.\;\, \con{f}(n) \teq \con{f}(n) +
1$, %ct and it is possible to construct consistent examples that do not meet it.
as well as other example of consistent definition.

@@

Another useful fragment is the class of \emph{productive} corecursive
functions. Corecursive functions are functions to a coalgebraic datatype.
These functions can call themselves forever, without their being inconsistent.
Productive corecursive functions are functions that progressively reveal parts
of their (potentially infinite)
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by $\con{scons}$,
the definition
$\{\forallf{\const{e}} n.\;\, \const{e}(n) \teq \con{scons}(n,\; \const{e}(n + 1))\}$
falls within this fragment, since each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specify the functions it defines.
It is even possible to mix recursion and corecursion in the same function
\cite{blanchette-et-al-2015-fouco}.

\begin{definition}
Let $\DDD = \{ \forallf{\con{f}_1} \vec x.\; \varphi_1, \ldots,
\forallf{\con{f}_n} \vec x.\; \varphi_n \}$ be a set of function definitions.
This set forms a \emph{layered }(\emph{co})\emph{recursive specification} if
$\const{f}_i$'s definition may not syntactically refer to $\const{f}_{j}$ for
all $i < j$ and each $\const{f}_i$ is either well-founded recursive
or productive corecursive.
\end{definition}

\begin{definition}\rm
Let $\DDD$ be a set of function definitions for $\sfundefs{\Sigma}$. The set
$\DDD$ is \emph{layered-total} if for every
$(\stypes{\Sigma},\, \sfuns{\Sigma} - \sfundefs{\Sigma})$-interpretations
$\J_0$, there exists a unique
$\Sigma$-interpretation $\J$ that extends $\J_0$
and for which $\J \models \DDD$.
\end{definition}

\begin{theorem}
If
$\DDD$ is a total set of function definitions for
$\sfundefs{\Sigma} = \{\con{f}_1,\ldots,\con{f}_n\}$, then it is admissible.
\end{theorem}
\begin{proofsketch}
Let $\psi$ be a satisfiable formula closed under function expansion with
respect to $\DDD$. We show that $\psi \wedge \bigwedge \Delta$ is also
satisfiable. Let $\I$ be a model of $\psi$, and
let $\J_0$ be the restriction of $\I$ to the function symbols in $\sfuns{\Sigma} -
\sfundefs{\Sigma}$. By hypothesis, there exists a $\Sigma$-interpretation
$\J$ such that $\J \models \Delta$.

@@

>>>>

\begin{abstract}
Model finding in SMT solvers is
generally limited to universally quantified formulas in some restricted fragment.
This paper introduces a translation which reduces axioms specifying a large
class of recursive functions, including well-founded (terminating),
to universally quantified formulas for which known
techniques are applicable.
An empirical evaluation confirms that the approach improves 
the performance of existing solvers on benchmarks from two sources. 
The translation is implemented as a preprocessor in the solver CVC4.
\end{abstract}

@@

In this paper, we present a method for translating formulas involving recursice function
definitions into formulas where finite model finding techniques can be applied.
The recursive functions must meet a semantic criterion to be admissible
(Section~\ref{sec:prelim}). This criterion is met by well-founded (terminating)
recursive function definitions. It is not met by inconsistent
definitions such as $\forall n : {\typeint}.\;\, \con{f}(n) \teq \con{f}(n) +
1$. %ct and it is possible to construct consistent examples that do not meet it.
%as well as other examples of consistent definition.

@@

\begin{theorem}\ref{thm:adm}
If
$\DDD$ is a set of well-founded function definitions for
$\sfundefs{\Sigma} = \{\con{f}_1,\ldots,\con{f}_n\}$, then it is admissible.
\end{theorem}
\begin{proofsketch}
Let $\psi$ be a satisfiable formula closed under function expansion with
respect to $\DDD$. We show that $\psi \wedge \bigwedge \Delta$ is also
satisfiable. Let $\I$ be a model of $\psi$, and
let $\J_0$ be the restriction of $\I$ to the function symbols in $\sfuns{\Sigma} -
\sfundefs{\Sigma}$. By hypothesis, there exists a $\Sigma$-interpretation
$\J$ such that $\J \models \Delta$.

@@

Another useful fragment is the class of \emph{productive} corecursive
functions. Corecursive functions are functions to a coalgebraic datatype.
These functions can call themselves forever, without their being inconsistent.
Productive corecursive functions are functions that progressively reveal parts
of their (potentially infinite)
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by $\con{scons}$,
the definition
$\{\forallf{\const{e}} n.\;\, \const{e}(n) \teq \con{scons}(n,\; \const{e}(n + 1))\}$
falls within this fragment, since each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specify the functions it defines.
It is even possible to mix recursion and corecursion in the same function
\cite{blanchette-et-al-2015-fouco}. We believe Theorem~\ref{thm:adm} can
be extended to such specifications.
